Bounds, Semantics/Pragmatics-Divide, and Evolution

### 2016-06-21 paper/00specification.tex
clean up, observations by sampling and parametrized learning for single scalar pair

\section{Short exposition}
The plan is to expand our work on the semantics/pragmatics divide by means of a (more extended) study on the lack of upper-bounds in the literal meaning of scalar expressions. The main idea of this extension, other than providing a more detailed treatment as well as a comparison with iterated learning, is to look at the effect multiple scalar pairs have on the dynamics. That is, we consider larger language fragments consisting of more than one pair of expressions of the ``some but not all'' vs. ``all''-form.

\section{Differences to CogSci setup.}
\subparagraph{(I) Lexica and types.} $3$ pairs of scalar expressions. Each pair is a $(2,2)$-Boolean matrix, which means that a lexicon is now a $(8,8)$-matrix. We have two types of signaling behavior, literal and Gricean, and $6$ different $(2,2)$-lexica. The latter is the set of possible semantics of a single scalar pair. With three scalar pairs we have $6 \cdot 6 \cdot 6 = 216$ $(8,8)$-lexica. $216$ times two signaling behaviors makes for a total of $432$ types.

\subparagraph{(II) Observations.} Before, the set of all production observations was $O =$\linebreak  $\{\tuple{\tuple{s_1,m_i},\tuple{s_2,m_j}} | m_i, m_j \in M\}$. A member of $O$ encodes that a teacher produced $m_i$ in state $s_1$ and $m_j$ in $s_2$, i.e. it encodes one witnessed message for each state. A datum $d$ was a sequence of length $k$ of members of $O$. Learners witnessed such data sequences. Now, more in line with \citet{griffiths+kalish:2007}, $O = \{\tuple{s_i,m_j} | s_i \in S, m_j \in M\}$ and $d$ is a sequence of length $k$ of members of $O$. The main difference is that now some $d$ do not provide any production information for some states.

\subparagraph{(III) Computational load.} Computing the $Q$-matrix from the set of all observation sequences of length $k$ is computationally intractable for large $k$. Here, we follow \citet{griffiths+kalish:2007} as well in sampling $1000$ sequences $d$ from the set of all sequences when $k > 4$. We use this sampled subset to compute $Q$ instead of the full set of sequences\footnote{Thomas: Assuming this solves the problem. My computer can't handle the computation even with sampling but a better computer may. I'm still waiting on people from my project to address my request for computation time on our server.}

\subparagraph{(IV) Mutation/Learning matrix.} If the results warrant it, classical (unweighted) learning instead of communal (weighted) learning.\footnote{Thomas: We can save the results for both cases in one go and decide on this later.}

\paragraph{Further modifications we may want to keep in mind (but currently don't make use of).}
\subparagraph{(I) Cost for pragmatic reasoning.} At least in the CogSci setup the effect of adding cost to pragmatic reasoning is unsurprising: High cost for pragmatic signaling lowers the prevalence of pragmatic types. Lexica that semantically encode an upper-bound benefit the most from this. However, the cost needed to be substantial to make the pragmatic English-like lexicon stop being the incumbent type (particularly when learning is communal).

@@

@@

We consider a population of players with two signaling strategies, literal and Gricean (level $0$ and $1$ below), each equipped with one of $216$ lexicons. This yields a total of $432$ distinct player types $t \in T$. However, each lexicon is composed of three independent lexical pairs. That is, while $|M| = |S| = 8$, we can decompose each lexicon into three independent $(2,2)$-matrices. A lexicon is a combination of three such matrices. These are listed in Table \ref{tab:lexica}.

@@

$L_4$ = $\begin{pmatrix} 0 & 1 \\ 1 & 0 \end{pmatrix}$ &
$L_5$ = $\begin{pmatrix} 0 & 1 \\ 1 & 1 \end{pmatrix}$ &
$L_6$ = $\begin{pmatrix} 1 & 1 \\ 1 & 0 \end{pmatrix}$
\end{tabular}
\caption{{\footnotesize Set of possible $(2,2)$-matrices, i.e. possible semantics for a pair of scalar expressions.}}
\label{tab:lexica}
\end{table}

@@

\subsection{Part 2: Mutation \& replication.}
The proportion of players of type $i$, $x_i$, is initialized as an arbitrary distribution over $T$. $p^\star \in \Delta(T)$ is learning a prior over (player) types dependent only on the lexicon of the type. 
\begin{itemize}
    \item $f_i = \sum_j x_j U(x_i,x_j)$
    \item $\Phi = \sum_i x_i f_i$
    \item $Q_{ij} \propto \sum_d P(d|t_i) \; P(t_j|d)$, where $P(t_j|d) \propto p^\star(t_j) P(d|t_j)$ and $d$ is a sequence of observations of length $k$ of the form \tuple{\tuple{s_i,m_j}, ... \tuple{s_k, m_l}}.
\end{itemize}

\paragraph{Weighted (communal) learning vs. unweighted (parental) learning}

\begin{itemize}
	\item For parental learning (standard RMD): $\dot x_i = \sum_j Q_{ji} \frac{x_j f_j}{\Phi}$
	\item For communal learning (our variant in CogSci): $\dot x_i = \sum_j N_{ji} \frac{x_j f_j}{\Phi}$, where $N_{ji} = x'_{i} Q_{ji}$ and $x'_i$ is the proportion of type $i$ in the previous generation.
\end{itemize}

\subsection{Model parameters \& procedure} 
\begin{enumerate}
  \item Sequence length $k$
  \item Pragmatic production parameter $\alpha$
  \item Rationality parameter $\lambda$
  \item Learning prior over types (lexica); cost parameter $c$. $p^\star(t_i) \propto n - c \cdot r$ where $n$ is the total number of states and $r$ that of upper-bounded messages only true of $s_1$ in $t_i$'s lexicon (if only $s_1$ is true of a message, then this message encodes an upper-bound). Then the score for $L_1$, $L_3$, $L_5$ is $2$, that of $L_4$ and $L_6$ is $2-c$, and that of $L_2$ is $2-2c$; Normalization over lexica scores yields the prior over lexica (which is equal to the prior over types).   
  \item Prior over meanings ($pr$). We assume that $pr(s) = \frac{1}{|S|}$ for all $s$.
  \item True state distribution ($P$). We currently assume that $P = \frac{1}{|S|}$ but it may be interesting to vary this
\end{enumerate}

\paragraph{Procedural description.} The game is initialized with some arbitrary distribution over player types. At the game's onset we compute $Q$ once based on the set of sequences $D$. If $k > 4$, $1000$ sequences are sampled from $D$ and used to compute $Q$ (instead of the full set). Replicator dynamics are computed based on the fitness of each type in the current population as usual. In the case of community learning, $Q$ is weighted by the previous population vector in each new generation.

\paragraph{Presentation.} We have a large number of types, some of which are highly similar. For instance, a type that uses $L_4$, $L_5$ and $L_4$ for her respective three scalar pairs is a different type than one that uses $L_5$, $L_4$ and $L_4$. As our goal is to elucidate the role and prevalence of upper-bounds, we may summarize the results not by type but by lexica counts. For instance, if the previous two types where all we had this would mean that $L_4$'s proportion is roughly $.66$ and $L_5$'s $.33$ (irrespective of signaling behavior). An example is given below for communal learning with a single scalar pair against three scalar pairs ($\alpha = 1, c = .7, \lambda = 30, k = 3$):

\begin{figure}[h] %htb for exact position
	\centering
	\resizebox{13cm}{!}{\includegraphics{../plots/wgh-single-vs-multi-k3-c70}}
\end{figure}

>>>>

\section{Short exposition}
The plan is to expand our work on the semantics/pragmatics divide by means of a more involved \& principled study on the lack of upper-bounds in the literal meaning of scalar expressions. The main idea of this extension, other than providing a more detailed treatment as well as a comparison with iterated learning, is to do away with the assumption of communal learning. Potential extensions include more scalar pairs, the addition of an inductive bias towards uniform lexica across scalar pairs, and a variation of state frequencies.

\section{Differences to CogSci setup.}
\subparagraph{(I) No communal learning.} Only parental learning is considered

\subparagraph{(II) Sequences and atomic observations.} Before, the set of all observations was $O =$\linebreak  $\{\tuple{\tuple{s_1,m_i},\tuple{s_2,m_j}} | m_i, m_j \in M\}$. A member of $O$ encodes that a teacher produced $m_i$ in state $s_1$ and $m_j$ in $s_2$, i.e., it encodes one witnessed message for each state. A datum $d$ was a sequence of length $k$ of members of $O$. Learners witnessed such data sequences. Now, more in line with \citet{griffiths+kalish:2007}, $O = \{\tuple{s_i,m_j} | s_i \in S, m_j \in M\}$ and $d$ is a sequence of length $k$ of members of $O$. The main difference is that now some $d$ do not provide any production information for some states.

\subparagraph{(III) Observations as production.} Instead of taking the space of all possible sequences of length $k$ into consideration, we take sample from $O$ $k$-times according to the production probabilities of each type; $P(o = \tuple{s,m} | t_i) = P(s) P(m|s,t_i)$. $n$ such $k$-length sequences are sampled for each type. As a consequence, the data used for computing $Q_i$ is not the same as that used for $j$ $(i \neq j)$.

\subparagraph{(IV) Parametrized learning} $Q_{ij} \propto [\sum_d P(d|t_i) P(t_j|d)]^l$, where $l =1$ corresponds to probability matching and, as $l$ increases towards infinity, to MAP.

\section{Further modifications we may want to keep in mind (but currently don't make use of)}
\subparagraph{(I) Cost for pragmatic reasoning.} At least in the CogSci setup the effect of adding cost to pragmatic reasoning is unsurprising: High cost for pragmatic signaling lowers the prevalence of pragmatic types. Lexica that semantically encode an upper-bound benefit the most from this. However, the cost needed to be substantial to make the pragmatic English-like lexicon stop being the incumbent type (particularly when learning is communal).

@@

\subparagraph{(V) More scalar pairs.} Taking into consideration more than one scalar pair. Preliminary results suggest that this does not influence the results in any meaningful way without further additions, e.g. by (III).

\subparagraph{(VI) More lexica.} Not necessary. Preliminary results suggest that considering more lexica has no noteworthy effect on the dynamics (tested with all possible 2x2 lexica).

\subparagraph{(VII) State frequencies.} Variations on state frequencies. This may have an interesting interaction with (III).

\subparagraph{(VIII) Reintroduction of communal learning.} One possibility: The probably $N_{ij}$ with which a child of $t_i$ adopts $t_j$ could be the weighted sum of $Q_{ij}$ (as before) and a vector we get from learning from all of the population: $L_j = \sum_d P(d | \vec{p})  P(t_j | d)$, where $P(d | \vec{p}) = \sum_{i} P(d | t_i)  \vec{p}_i$ is the probability of observing $d$ when learning from a random member of the present population distribution.

@@

We consider a population of players with two signaling behaviors, literal and Gricean (level $0$ and $1$ below), each equipped with one of $6$ lexicons. This yields a total of $12$ distinct player types $t \in T$. $|M| = |S| = 2$, i.e., a lexicon is a $(2,2)$-matrix. These are listed in Table \ref{tab:lexica}.

@@

$L_4$ = $\begin{pmatrix} 0 & 1 \\ 1 & 0 \end{pmatrix}$ &
$L_5$ = $\begin{pmatrix} 0 & 1 \\ 1 & 1 \end{pmatrix}$ &
$L_6$ = $\begin{pmatrix} 1 & 1 \\ 1 & 0 \end{pmatrix}$
\end{tabular}
\caption{{\footnotesize Set of possible $(2,2)$-matrices, i.e., a lexicon.}}
\label{tab:lexica}
\end{table}

@@

\subsection{Part 2: Mutation \& replication.}
The proportion of players of type $i$, $x_i$, is initialized as an arbitrary distribution over $T$. $p^\star \in \Delta(T)$ is learning a prior over (player) types dependent only on the lexicon of the type. 
\begin{itemize}
    \item $f_i = \sum_j x_j U(x_i,x_j)$
    \item $\Phi = \sum_i x_i f_i$
    \item $Q_{ij} \propto [\sum_d P(d|t_i) \; P(t_j|d)]^l$, where $P(t_j|d) \propto p^\star(t_j) P(d|t_j)$, $d$ is a sequence of observations of length $k$ of the form \tuple{\tuple{s_i,m_j}, ... \tuple{s_k, m_l}}, and $l \geq 1$ is a learning parameter.
	\item For parental learning (standard RMD): $\dot x_i = \sum_j Q_{ji} \frac{x_j f_j}{\Phi}$
\end{itemize}

\subsection{Model parameters \& procedure} 
\begin{enumerate}
  \item Sequence length $k$
  \item Pragmatic production parameter $\alpha$
  \item Rationality parameter $\lambda$
  \item Learning prior over types (lexica); cost parameter $c$. $p^\star(t_i) \propto n - c \cdot r$ where $n$ is the total number of states and $r$ that of upper-bounded messages only true of $s_1$ in $t_i$'s lexicon (if only $s_1$ is true of a message, then this message encodes an upper-bound). Then the score for $L_1$, $L_3$, $L_5$ is $2$, that of $L_4$ and $L_6$ is $2-c$, and that of $L_2$ is $2-2c$; Normalization over lexica scores yields the prior over lexica (which is equal to the prior over types).   
  \item Prior over meanings ($pr$). We assume that $pr(s) = \frac{1}{|S|}$ for all $s$.
  \item True state distribution ($P$). We currently assume that $P = \frac{1}{|S|}$ but it may be interesting to vary this
  \item Learning parameter $l \geq 1$ with $1$ corresponding to probability matching, and MAP as $l$ approaches infinity
  \item $n$ is the sample of sequences of observations of length $k$ sampled from the production probabilities of each type
  \item Number of generations $g$
\end{enumerate}

\paragraph{Procedural description.} The game is initialized with some arbitrary distribution over player types. At the game's onset we compute $Q$ once based on the sets  of sequences $D$ (one for each parent type). Replicator dynamics are computed based on the fitness of each type in the current population as usual. $Q$ is computed anew for each independent run (of $g$ generations) given that it depends on $D$, which is sampled from production probabilities.
